---
title: "Binary Search Tree"
output:
  html_document: 
    toc: yes
    toc_float: yes
    highlight: haddock
css: ./assets/css/styles.css
# knit: pagedown::chrome_print
bibliography: ./assets/citation/csc112.bib
csl: ./assets/citation/institute-of-mathematical-statistics.csl 
link-citations: yes
nocite: '@*'
---


```{r include=FALSE}
library(knitr)
library(graphics)
library(drc)
library(data.tree)
library(DiagrammeR)
library(kableExtra)
library(rmarkdown)
knitr::opts_chunk$set(echo = FALSE, out.width = 600, eval = TRUE, message = FALSE, warning = FALSE, fig.align = "center")

```


## Binary Nodes

<span class="sp4">a.</span> Every node can have up to two children: a left child and a right child

```{r include=FALSE, fig.align='center'}
header <- Node$new("Header")
root1 <- Node$new("Root")
child <- Node$new("LeftChild")
sibling <- Node$new("RightChild")

header$AddChildNode(root1)
root1$AddChildNode(child)$AddSiblingNode(sibling)

SetNodeStyle(header, style = "filled", shape="rectangle",fillcolor = "aliceblue", fontname = "helvetica", fontsize=12, position="center")
SetEdgeStyle(header, arrowhead = "vee", color = "grey35")
SetNodeStyle(root1, style = "filled", shape = "circle", fillcolor = "thistle", fontname = "helvetica", fontsize=6)
SetNodeStyle(root1$LeftChild, inherit = FALSE, fillcolor = "lightseagreen", fontcolor = "black", fontsize=5)
SetNodeStyle(root1$RightChild, inherit = FALSE, fillcolor = "honeydew3", fontcolor = "black", fontsize=4.5)

```



```{r, out.height=200, fig.align="center", fig.show='hold'}
library(networkD3)
acmeNetwork <- ToDataFrameNetwork(header, "name")
                                                    
simpleNetwork(acmeNetwork[-3], fontSize = 12)
```




### Types of Binary Nodes

<div class="a3">

a. **Root:** the first node

b. **Leaf:** a node with no children 

c. **Interior Nodes:** nodes that are neither the root or a leaf 

</div>

\ 


------------------------------------------------


## Traversals of Binary Trees

### Mechanisms

<span class="sp4">a.</span> In general, there isn't an explicit order for traversing binary tree arrangements. Here are three example mechanisms:


<div class="a4">

<span class="sp14">1.</span> Pre-order traversal 

In this traversal method, the program visits each node before visiting any child nodes. Usually, it goes to a node first, then each node in its left subtree, and then each node in its right subtree.

<span class="sp14">2.</span> $\bigstar$ In-order traversal

In this traversal method, the program first traverses the left subtree, then visits the root, and then traverses the right subtree.



<span class="sp14">3.</span> Post-order traversal 

In this traversal method, the program first traverses the left subtree, then the right subtree, and lastly, visits the root node.


</div>

\ 

#### Summary:

Traversing a binary tree requires creating an iterator that traverses all subtrees. 

- we perform implementations of iterators for each subtree traversal


------------------------------------------------

## Binary Search Tree

### Definition

A binary search tree is a binary tree structure with the following attributes: 

<span class="lp4">a.</span> The left subtree of a node holds only nodes with values lesser than the node


<span class="lp4">b.</span> The right subtree of a node has only nodes with values greater than the node

<span class="lp4">c.</span> Both the left and right subtrees must also be a binary search tree

\ 

<div class="a1">

<span class="lp10">a.</span> unlike the BinaryTree, the BinarySearchTree provides one iterator method: an in-order traversal


<span class="lp10">b.</span> when implementing a BinarySearchTree, we're implementing an OrderedStructure: methods that accept and return values compared to one another

- we assume the data is Comparable, and the natural order is sufficient

<span class="lp10">c.</span> we can use alternative Comparators for elements that do not directly implement compareTo


<span class="lp10">d.</span> only methods depended on are the compatibility of the Comparator and the tree elements


</div>



\ 

###  Example. Complete Binary Search Tree with 4 Levels.


<span class="sp4">a.</span> The below is an example of a **complete** binary tree:


```{r BinaryTree, include=FALSE}
root <- Node$new("b50")
child <- Node$new("a50")
sibling <- Node$new("c50")
root$AddChildNode(child)$AddSiblingNode(sibling)

child2 <- Node$new("a40")
sibling2 <- Node$new("a60")
child$AddChildNode(child2)$AddSiblingNode(sibling2)

child3 <- Node$new("c40")
sibling3 <- Node$new("c60")
sibling$AddChildNode(child3)$AddSiblingNode(sibling3)

child4 <- Node$new("a35")
sibling4 <- Node$new("a45")
child2$AddChildNode(child4)$AddSiblingNode(sibling4)

child5 <- Node$new("a55")
sibling5 <- Node$new("a65")
sibling2$AddChildNode(child5)$AddSiblingNode(sibling5)

child6 <- Node$new("c35")
sibling6 <- Node$new("c45")
child3$AddChildNode(child6)$AddSiblingNode(sibling6)

child7 <- Node$new("c55")
sibling7 <- Node$new("c65")
sibling3$AddChildNode(child7)$AddSiblingNode(sibling7)
```



```{r, include=FALSE}
SetEdgeStyle(root, arrowhead = "vee", color = "grey35", penwidth = 2, position = "center")
SetNodeStyle(root, style = "filled", shape = "circle", fillcolor = "thistle", 
            fontname = "helvetica", tooltip = GetDefaultTooltip, position = "center", width=0.5, fontsize=16)
SetNodeStyle(root$a50, inherit = FALSE, fillcolor = "#ECD999", fontcolor = "black")
SetNodeStyle(root$c50, inherit = FALSE, fillcolor = "#ECD999", fontcolor = "black")
SetNodeStyle(child$a40, inherit = FALSE, fillcolor = "#ffa089", fontcolor = "black")
SetNodeStyle(child$a60, inherit = FALSE, fillcolor = "#ffa089", fontcolor = "black")
SetNodeStyle(sibling$c40, inherit = FALSE, fillcolor = "#5CBD92", fontcolor = "black")
SetNodeStyle(sibling$c60, inherit = FALSE, fillcolor = "#5CBD92", fontcolor = "black")
SetNodeStyle(child2$a35, inherit = FALSE, fillcolor = "#EF4868", fontcolor = "black")
SetNodeStyle(child2$a45, inherit = FALSE, fillcolor = "#EF4868", fontcolor = "black")
SetNodeStyle(sibling2$a55, inherit = FALSE, fillcolor = "#FBA0E3", fontcolor = "black")
SetNodeStyle(sibling2$a65, inherit = FALSE, fillcolor = "#FBA0E3", fontcolor = "black")
SetNodeStyle(child3$c35, inherit = FALSE, fillcolor = "#38BDBB", fontcolor = "black")
SetNodeStyle(child3$c45, inherit = FALSE, fillcolor = "#38BDBB", fontcolor = "black")
SetNodeStyle(sibling3$c55, inherit = FALSE, fillcolor = "#9bddff", fontcolor = "black")
SetNodeStyle(sibling3$c65, inherit = FALSE, fillcolor = "#9bddff", fontcolor = "black")
```


```{r, out.height=230, fig.align="center", fig.show='hold'}
graphics::plot(root)
```


#### Summary:

<div class="a3">

<span class="lp10">a.</span> `Levels:` The root and each interior node have two children --a right and left child; hence, there are $4$ **full** levels.

<span class="lp10">b.</span> `Nodes:` The number of nodes can be calculated by $2^{n}-1$ where $n$ represents the number of full levels; hence, there are $15$ nodes.

</div>

\ 

-----------------


### Types of Binary Search Trees

<span class="sp3">I.</span> Complete Binary Trees

<span class="lp3">a.</span> Complete Binary Trees give excellent time complexities of $\mathbb{\bf O}\left(\log2\left(n\right)\right)$, which helps for find and insert methods in a Java data structure. 

<span class="lp3">b.</span> The number of **nodes** can be calculated with, $\mathbb{\bf n}\text{ full levels:}\left(2^\mathbb{\bf n}-1\right) \text{ nodes}$.




```{r, out.height=160, fig.align="center", fig.show='hold'}
graphics::plot(root1)
```


\ 

<span class="sp3">II.</span> "Real" Binary Trees

<span class="lp3">a.</span> It's possible to have a non-complete binary tree, which means not every parent node has two children nodes. 

<span class="lp3">b.</span> In the worst case, our binary tree could be a LinkedList with $\mathbb{\bf O}\left(n\right)$ complexities.



\ 


------------------------------------------------


## Java Code.

### Binary Search Tree

<div class="b1">

<span class="sp4">I.</span> Main.java

Main method to demonstrate binary tree and binary tree node

<span class="sp4">II.</span> BinaryNode.java


Simple BinaryNode method using a generic that extends Comparable

<span class="sp4">III.</span> BinaryTree.java

Simple BinaryTree method using a generic E that extends Comparable

</div>

\ 

---------------------------------

### 1. Main.java

#### Outline 

<div class="b1">

<span class="lp6">a.</span> creates an empty `BinaryTree` 

<span class="lp6">b.</span> creates random integers to insert into the tree

<span class="lp6">c.</span> prints out the tree's nodes and order 

<span class="lp6">d.</span> prints out the tree

</div>


\ 

 


```java
import java.util.concurrent.ThreadLocalRandom;
import java.util.Scanner;

public class Main {
  public static void main(String[] args) {
    
    // create an empty BinaryTree
    BinaryTree<Integer> myTree = new
    BinaryTree<Integer>();
    
    // instantiate new scanner
    Scanner keyboard = new Scanner(System.in);
    
    // create random integers and insert them into myTree
    for (int i=0; i< 5; i++){
      
      // generate random number in range [-15, 50]
      System.out.print("Enter item #"+i+": ");
      int randValue = keyboard.nextInt();
      
      // insert each number into myTree
      myTree.insert(randValue); }
    
    // System.out.println(myTree);
    
    System.out.print("The nodes of tree are (inorder): ");
    System.out.println(myTree.inorder());
    System.out.print("The breadth order of the tree is: ");
    System.out.println(myTree);
  }
}
```



---------------------------------

### 2. BinaryNode.java

#### Outline

<div class="b1">

<span class="lp10">a.</span> uses a generic that extends Comparable `.compareTo()`

<span class="lp10">b.</span> each node contains three objects: data, a left child, and a right child

<span class="lp10">c.</span> alternate constructor creates a new leaf in binary tree and getters return objects

<span class="lp10">d.</span> `insert(E item)` method recursively inserts item as a new leaf

<span class="lp10">e.</span> `boolean` method checks if the Left or Right child is empty

<span class="lp10">f.</span>  `inorder` method recursively traverses the nodes

- if there is a nonempty left child, check there first
- if there is a nonempty right child, check there last

</div>


\ 

 


```java
public class BinaryNode<E extends Comparable> {
  
  // each node contains three objects
  private E  Data;
  private BinaryNode<E> leftchild;
  private BinaryNode<E> rightchild;

  // alternate constructor: creates a new leaf in binary tree
  public BinaryNode(E item){
    Data = item;
    leftchild = null;
    rightchild = null; }
  
  // getters: return objects
  public E getData(){
    return Data; }
  public BinaryNode<E> getLeft(){
    return leftchild; }
  public BinaryNode<E> getRight(){
    return rightchild; }
  
  //  recursively insert item as a new leaf
  public void insert(E item){
    int compareV = item.compareTo(Data);
    
    // ignore items already in tree
    if (compareV ==0) { 
      return; }
    
    // left child
    else if (compareV < 0){
      if (leftchild == null){
        leftchild = new BinaryNode<E>(item); }
      else {
        leftchild.insert(item); }}
    
    // right child
    else {
      if (rightchild == null){
        rightchild = new BinaryNode<E>(item);
        return; }
      else {
        rightchild.insert(item); }}
  }
  
  // boolean checks whether Left or Right child is empty
  public boolean isLeftChildEmpty(){
    return leftchild == null; }
  public boolean isRightChildEmpty(){
    return rightchild == null; }
  
  // recursive inorder traversal of the nodes
  public String inorder(){
    String Result = " ";
    
    // if there is a nonempty left child, check there first
    if (!isLeftChildEmpty()){
      Result = Result + leftchild.inorder(); }
    // now print value at the node
    Result = Result + " " + Data;
    
    // if there is a nonempty right child, check there last
    if (!isRightChildEmpty()){
      Result = Result + " " + rightchild.inorder(); }
    return Result; }
  }
```

---------------------------------


### 3. BinaryTree.java

#### Outline


<div class="b1">

<span class="lp7">a.</span> uses a generic E to extend Comparable `.compareTo()`

<span class="lp7">b.</span> default constructor establishes tree's root node

<span class="lp7">c.</span> `insert(E item)` method inserts an item into the tree as a binary node

- if tree is empty, create a binary node as the root containing an item
- insert item as a node using the order relation

<span class="lp7">d.</span> `inorder` method prints the tree in order

- calls recursive `BinaryNode inorder()` and returns roots
    
<span class="lp7">e.</span> `ToString` method prints tree in a breadth first order

</div>


\ 

 


```java
import java.util.LinkedList;
import java.util.Queue;

public class BinaryTree<E extends Comparable> {
  
  private BinaryNode<E> root;
  
  // default constructor
  public BinaryTree(){
    root = null; }
  
  
  // insert an item into the tree as a binary node
  public void insert(E item) {
    
    // if tree is empty, create a binary node as the root containing an item
    if (root == null) {
      root = new BinaryNode<E>(item);
      return; }
    
    // insert item as a node using the order relation
    root.insert(item); }
  
  
  // print tree in inorder
  public String inorder(){
    String Result = "";
    if (root == null){
      return Result; }
    
    // call recursive BinaryNode inorder()
    return root.inorder(); }
  

  // print tree in a breadth first order
  public String toString(){
    Queue<BinaryNode<E>> myQueue = new LinkedList<BinaryNode<E>>();
    String Result = "";
    
    if (root == null) return Result;
    myQueue.add(root);
    
    while (!myQueue.isEmpty()){
      BinaryNode<E> current = myQueue.remove();
      Result = Result + " " + current.getData();
      if (!current.isLeftChildEmpty()){
        myQueue.add(current.getLeft()); }
      if (!current.isRightChildEmpty()){
        myQueue.add(current.getRight()); }}
    return Result; }
  
  }
```



\ 

---------------------------------


## Summary

\ 

<div class="a3">

The binary search tree requires an order on the nodes of the binary tree. 

- At each node in the search method, the program decides to go left or right.
- If the tree is short and relatively balanced, these decisions will eliminate many of the remaining candidate values.

\ 

The binary search tree is the result of inserting new values. 

- method puts each new data point on a leaf; hence, the internal nodes remain unchanged, thereby making the structure reasonably static
- insufficient data allocation will cause a degraded tree structure to negatively affect the method's performance

</div>


\ 

### Time Complexity

<div class="b1">

Each operation of a Binary Search Tree has a worst-case time complexity proportional to the tree's height.

**Note.** Checking leaves, adding leaves, or removing roots are some of the most time-consuming operations. Therefore, for logarithmic behavior, we want the tree to be as short as possible.


<span class="sp12">i.</span> If we insert values using descending order, the tree is left-skewed.


<span class="sp12">ii.</span> If we insert values using ascending order, the tree is right-skewed.

The program performs better if we shuffle the values beforehand:

<span class="lp12">i.</span> the tree becomes shorter and more balanced

<span class="lp12">ii.</span> the expected insertion time is $O(\log n)$




</div>

\ 

## Traversal Example

An InOrder traversal visits the nodes in the following order:


```{r, include=FALSE}
root <- Node$new("25")
child <- Node$new("15")
sibling <- Node$new("50")
root$AddChildNode(child)$AddSiblingNode(sibling)

child2 <- Node$new("10")
sibling2 <- Node$new("22")
child$AddChildNode(child2)$AddSiblingNode(sibling2)

child3 <- Node$new("35")
sibling3 <- Node$new("70")
sibling$AddChildNode(child3)$AddSiblingNode(sibling3)

child4 <- Node$new("4")
sibling4 <- Node$new("12")
child2$AddChildNode(child4)$AddSiblingNode(sibling4)

child5 <- Node$new("18")
sibling5 <- Node$new("24")
sibling2$AddChildNode(child5)$AddSiblingNode(sibling5)

child6 <- Node$new("31")
sibling6 <- Node$new("44")
child3$AddChildNode(child6)$AddSiblingNode(sibling6)

child7 <- Node$new("66")
sibling7 <- Node$new("90")
sibling3$AddChildNode(child7)$AddSiblingNode(sibling7)
```



```{r, include=FALSE}
SetEdgeStyle(root, arrowhead = "vee", color = "#708090", position = "center")
SetNodeStyle(root, style = "filled", shape = "circle", fillcolor = "#9bddff",fontcolor = "black", fontsize=12,
            fontname = "helvetica", tooltip = GetDefaultTooltip, position = "center", width=0.6)
```


```{r, out.height=210, fig.align="center", fig.show='hold'}
graphics::plot(root)
```




$$\begin{matrix}4 & \mathbf{10} & 12 & \mathbf{15} & 18 & \mathbf{22} & 24 & \Large\mathbf{25} & 31 & \mathbf{35} & 44 & \mathbf{50} & 66 & \mathbf{70} & 90\end{matrix}$$


\ 




- [Helpful resource](https://www.cs.usfca.edu/~galles/visualization/BST.html) for plotting nodes in binary search trees


------------------------


## References

